Estimating Neighborhood Rents with Scraped Data

### 2018-07-03 dashboard/index.rmd
committing work through 7/3: the data are now full, i.e. Q1 2017 through end of Q2 2018. The manuscript .Rmds are effectively synced (index.rmd = manuscript.rmd with a different header). Also, cleaned up the helpers so all apply the same data load procedure

#establish object for modeling based off of adjusting `panel`
sea_df <- panel %>% 
  mutate(actualRent = med1B, #first preserve the observed / observed-to-date
         med1B = ifelse(listingQtr == as.yearqtr(Sys.Date()), NA, med1B)) %>% #set to NA for INLA to forecast current qtr
  mutate(p1B = n1B/nListings) %>% #compute share of 1B listings
  arrange(listingQtr, GISJOIN) #arange qtr, tract

@@

#table of fit statistics
fit_stats <- sea_df %>%
  mutate(train_test = ifelse(listingQtr == as.yearqtr(Sys.Date()), "Test", "Training")) %>%
  mutate(int_err = actualRent - int_Med,
         ns_err = actualRent - ns_Med,
         nsar1_err = actualRent - nsar1_Med,
         bym_err = actualRent - bym_Med,
         spt_err = actualRent - spt_Med) %>% 
  group_by(train_test) %>%
  summarize(int_rmse = rmse(int_err[!is.na(int_err)]),
            int_mae = mae(int_err[!is.na(int_err)]),
            ns_rmse = rmse(ns_err[!is.na(ns_err)]),
            ns_mae = mae(ns_err[!is.na(ns_err)]),
            nsar1_rmse = rmse(nsar1_err[!is.na(nsar1_err)]),
            nsar1_mae = mae(nsar1_err[!is.na(nsar1_err)]),
            bym_rmse = rmse(bym_err[!is.na(bym_err)]),
            bym_mae = mae(bym_err[!is.na(bym_err)]),
            spt_rmse = rmse(spt_err[!is.na(spt_err)]),
            spt_mae = mae(spt_err[!is.na(spt_err)])) %>%
  dplyr::select(train_test, ends_with("_rmse"), ends_with("_mae"))
```

@@

The goal of this research is to a.) understand the temporal dynamics of rent estimates in Seattle and b.) forecast the current quarter's rent levels based off of the prior periods. The focal series of models regress median one-bedroom rent asked values on different specifications of the panel's correlation structure (i.e. temporal and spatial). All of the candidate models' posterior distributions are estimated with [integrated nested Laplace approximations (INLA)](https://folk.ntnu.no/hrue/r-inla.org/papers/inla-rss.pdf) using the default, weakly-informative priors for all model hyperparameters. Throughout the following analyses, the training data are `r min(sea_df$listingQtr)` up to the prior quarter (i.e. `r as.yearqtr(Sys.Date())-.25`). The test period is a forecast for the current period and includes comparison to the appropriate median rent estimates for data observed so far in this period.

@@

```{r index.maps, echo=F}
obs <- sea_f %>%
  mutate(rent = actualRent) %>%
  filter(listingQtr == as.yearqtr(Sys.Date())) %>%
  mutate(var = paste0(as.yearqtr(Sys.Date()), "-to-Date")) %>%
  dplyr::select(GISJOIN, long, lat, group, rent, var, actualRent, bym_Med, NAMELSAD10, n1B)

pred <- sea_f %>%
  mutate(rent = bym_Med) %>%
  filter(listingQtr == as.yearqtr(Sys.Date())) %>%
  mutate(var = paste0(as.yearqtr(Sys.Date()), " Forecast")) %>%
  dplyr::select(GISJOIN, long, lat, group, rent, var, actualRent, bym_Med, NAMELSAD10, n1B)

comb <- bind_rows(obs, pred) %>%
  mutate(tooltip = paste0(NAMELSAD10, "<br>", "Forecast: $", round(bym_Med), "<br>", "Qtr-to-Date: $", round(actualRent), "<br>", "N Qtr-to-Date: ", n1B))
comb <- SharedData$new(comb, ~GISJOIN)

>>>>

#establish object for modeling based off of adjusting `panel`
sea_df <- panel %>% 
  mutate(actualRent = med1B, #first preserve the observed / observed-to-date
         med1B = ifelse(listingQtr == "2018 Q2", NA, med1B)) %>% #set to NA for INLA to forecast current qtr
  mutate(p1B = n1B/nListings) %>% #compute share of 1B listings
  arrange(listingQtr, GISJOIN) #arange qtr, tract

@@

#table of fit statistics
fit_stats <- sea_df %>%
  mutate(train_test = ifelse(listingQtr == "2018 Q2", "Test", "Training")) %>%
  mutate(int_err = actualRent - int_Med,
         ns_err = actualRent - ns_Med,
         nsar1_err = actualRent - nsar1_Med,
         bym_err = actualRent - bym_Med,
         spt_err = actualRent - spt_Med) %>% 
  group_by(train_test) %>%
  summarize(int_rmse = rmse(int_err[!is.na(int_err)]),
            int_mae = mae(int_err[!is.na(int_err)]),
            ns_rmse = rmse(ns_err[!is.na(ns_err)]),
            ns_mae = mae(ns_err[!is.na(ns_err)]),
            nsar1_rmse = rmse(nsar1_err[!is.na(nsar1_err)]),
            nsar1_mae = mae(nsar1_err[!is.na(nsar1_err)]),
            bym_rmse = rmse(bym_err[!is.na(bym_err)]),
            bym_mae = mae(bym_err[!is.na(bym_err)]),
            spt_rmse = rmse(spt_err[!is.na(spt_err)]),
            spt_mae = mae(spt_err[!is.na(spt_err)])) %>%
  dplyr::select(train_test, ends_with("_rmse"), ends_with("_mae"))
```

@@

The goal of this research is to a.) understand the temporal dynamics of rent estimates in Seattle and b.) forecast the current quarter's rent levels based off of the prior periods. The focal series of models regress median one-bedroom rent asked values on different specifications of the panel's correlation structure (i.e. temporal and spatial). All of the candidate models' posterior distributions are estimated with [integrated nested Laplace approximations (INLA)](https://folk.ntnu.no/hrue/r-inla.org/papers/inla-rss.pdf) using the default, weakly-informative priors for all model hyperparameters. Throughout the following analyses, the training data are 2017 Q1 up through 2018 Q1. The test period is a forecast for 2018 Q2 and includes comparison to the observed median rent estimates for data collected in this period.

@@

```{r index.maps, echo=F}
obs <- sea_f %>%
  mutate(rent = actualRent) %>%
  filter(listingQtr == "2018 Q2") %>%
  mutate(var = "2018 Q2 Observed") %>%
  dplyr::select(GISJOIN, long, lat, group, rent, var, actualRent, bym_Med, NAMELSAD10, n1B)

pred <- sea_f %>%
  mutate(rent = bym_Med) %>%
  filter(listingQtr == "2018 Q2") %>%
  mutate(var = "2018 Q2 Smoothed") %>%
  dplyr::select(GISJOIN, long, lat, group, rent, var, actualRent, bym_Med, NAMELSAD10, n1B)

comb <- bind_rows(obs, pred) %>%
  mutate(tooltip = paste0(NAMELSAD10, "<br>", "Smoothed: $", round(bym_Med), "<br>", "Observed: $", round(actualRent), "<br>", "N Observed: ", n1B))
comb <- SharedData$new(comb, ~GISJOIN)

### 2018-07-03 manuscript/index.Rmd
committing work through 7/3: the data are now full, i.e. Q1 2017 through end of Q2 2018. The manuscript .Rmds are effectively synced (index.rmd = manuscript.rmd with a different header). Also, cleaned up the helpers so all apply the same data load procedure

---
title: "Estimating Neighborhood Rents with Scraped Data"
author: "Chris Hess and Sarah Chasins"
date: "June 5, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    includes:
      in_header: "header.html"
      after_body: "../footer.html"
abstract: "Online classified ads are a key source of information about local rental housing markets. This research first assesses the agreement of census tract rent estimates based on scraped rental advertisements with existing data sources. The second component of the paper investivates temporal and spatial dynamics in one-bedroom rents among Seattle neighborhoods using spatial models. All models' posterior distribution are estimated with integrated nested Laplace approximations (INLA). Training data for models cover quarterly tract estimates for 2017 Q2 through 2018 Q1, while the test data are 2018 Q2 median values as observed through June 4th. The top model assumes that spatial and temporal dynamics are separable, i.e. independent of each other, and non-interactive. This spatial random-effects model with AR(1) temporal structure achieves a mean absolute error (MAE) of about $100 for the training data and $140 for the test data."
---

@@

Only the Craigslist and Dupre+Scott can offer insight into more recent median levels for the Seattle rental housing market. Both data suggest continued increases in the market-wide median rent for the first half of 2017, though the Craigslist slope is about twice in its size. Some of greater variation observed between the first and second data points of the Craigslist series may be related to missing listings (whether due to the early-stage process or outages), something which might bias the estimate for 2017 Q2 upward in favor of regularly posted commercial listings. Though the estimates here reflect deduplicated listings, regularly updating a listing raises its likelihood of being at the top of the index our scraper works off of, thereby increasing its selection probability. This volatility in Craigslist asking rents could also be brought on by recently-developed buildings in "lease-up" adding a glut of expensive listings to the quarters' samples.

However, by the final point of Dupre+Scott's series (Fall 2017, corresponding to September), the difference compared to the Craigslist median was about $75. Mirroring other accounts about a decline in rents during Winter 2018, the Craigslist data showed a fairly precipitous decline between

@@

Figure 5 shows a scatter plot/bubble plot of 2000-definition Census tract one-bedroom rent estimates between Craigslist and Dupre+Scott for the 75 tracts where comparable estimates exist. The purple dashed line indicates the loess smoother fit between the measures. Looking at the measure comparison, the first takeaway is that there is strong correlation ($r = .82$) and good agreement between the measures (mean absolute difference = $186). The Craigslist measure reflects asking rents and the Dupre+Scott measure reflects contract rents for a neighborhood (i.e. the asking rent less an incentives a landlord offers). Given the currently low vacancy rates in most properties, the only neighborhoods where there would theoretically be more salience to the asking vs contract rent distinction (i.e. a wider gap in these indicators) are at the high end of the distribution where demand for luxury apartments has slackened some. At the high-end of the neighborhood rent distribution, however, Craigslist and Dupre+Scott are in arguably the strongest agreement.

@@

An alternative explanation for the difference between measures comes in considering the units these properties are sampling from in a given neighborhood. The greater difference between the measures among lower-rent neighborhoods (with Craigslist estimating these as more expensive than Dupre+Scott) provides some evidence for this theory. These tracts are found towards the North and South edges of the city, and generally are more residential neighborhoods where 20+ unit apartment complexes are not the entire stock of rental unit properties. In these tracts, Craigslist has a non-zero probability of sampling from any larger building but also small houses, duplex/triplexes, mother-in-law suites and recently, [Accessory Dwelling Units (ADUs)](http://www.seattle.gov/DPD/Publications/CAM/cam116a.pdf){target="_blank"}.

To the extent that these smaller property units may tend to be newer, larger in housing unit square footage or offer added amenities, the true value of rent for the neighborhood will be higher than what it is estimated at using 20+ unit properties alone. Similarly, the more that smaller properties account for the overall population of rental housing units, the more likely that higher values indicated in Craigslist are not erroneous but instead reflective of a broader set of residential units.

@@

The following models learn from Craigslist data accumulated up till the prior quarter (i.e. `r as.yearqtr(Sys.Date())-1`) to accomplish a few different goals. The most prominent goal is to use the panel data's temporal and spatial structure to understand variation in rent levels over time and across space. A secondary goal is to forecast the Seattle neighborhood rent surface for the current quarter. Finally, the following models allow for relatively straightforward imputation of missing values based on time period and the rent values of proximate tracts.

We assume that the log of the median one-bedroom rent for Seattle neighborhoods is normally distributed and compute the posterior distributions of each model's hyperparameters and parameters $\theta_{it}$ using [integrated nested Laplace approximations (INLA)](https://folk.ntnu.no/hrue/r-inla.org/papers/inla-rss.pdf){target="_blank"} with the default, weakly-informative priors for all random-effects. We choose to log the dependent variable because all values are positive and this transformation improves model fit.

@@

The BYM Spatial with AR(1) is the [Besag-York-Mollie (1991)](https://link.springer.com/article/10.1007/BF00116466){target="_blank"} intrinsic conditional autoregressive spatial model with AR(1) and exchangeable hyperparameters for the panel's temporal structure (referred to as `bym` using shorthand). The random-effect $S_i$ is assumed to come from a normal distribution whose mean is the mean of the neighbors' random effects and variance proportional to the number of neighbors (formally, $v_i|v_j, j \in ne(i) ~ \mbox{Normal}(\bar{v_i}, \frac{\sigma^{2}_v}{m_j})$. Lastly, the BYM Spatiotemporal AR(1) (`spt`) adds an exchangeable random-effect ($\delta_i \sim Normal(0, \tau_\delta)$) for each period to allow for area-specific deviations in time trends.

@@

#establish object for modeling based off of adjusting `panel`
sea_df <- panel %>% 
  mutate(actualRent = med1B, #first preserve the observed / observed-to-date
         med1B = ifelse(listingQtr == as.yearqtr(Sys.Date()), NA, med1B)) %>% #set to NA for INLA to forecast current qtr
  mutate(p1B = n1B/nListings) %>% #compute share of 1B listings
  arrange(listingQtr, GISJOIN) #arange qtr, tract
```

@@

#spatial random effect model (BYM 1991) with AR(1)
form.bym <- log(med1B) ~ 1 +
  f(idtract, model = "bym", #ICAR spatial RE for tract neighbors + IID RE for tract
    scale.model = T,
    graph = "../output/seatract.graph") +
  f(idqtr, model = "ar1") +
  f(idqtr1, model = "iid")

@@

#table of fit statistics
fit_stats <- sea_df %>%
  mutate(train_test = ifelse(listingQtr == as.yearqtr(Sys.Date()), "Test", "Training")) %>%
  mutate(int_err = actualRent - int_Med,
         ns_err = actualRent - ns_Med,
         nsar1_err = actualRent - nsar1_Med,
         bym_err = actualRent - bym_Med,
         spt_err = actualRent - spt_Med) %>% 
  group_by(train_test) %>%
  summarize(int_rmse = rmse(int_err[!is.na(int_err)]),
            int_mae = mae(int_err[!is.na(int_err)]),
            ns_rmse = rmse(ns_err[!is.na(ns_err)]),
            ns_mae = mae(ns_err[!is.na(ns_err)]),
            nsar1_rmse = rmse(nsar1_err[!is.na(nsar1_err)]),
            nsar1_mae = mae(nsar1_err[!is.na(nsar1_err)]),
            bym_rmse = rmse(bym_err[!is.na(bym_err)]),
            bym_mae = mae(bym_err[!is.na(bym_err)]),
            spt_rmse = rmse(spt_err[!is.na(spt_err)]),
            spt_mae = mae(spt_err[!is.na(spt_err)])) %>%
  dplyr::select(train_test, ends_with("_rmse"), ends_with("_mae"))
```

The following tables show in-sample and out-of-sample (i.e. 2018 Q2) accuracy for the 5 model specifications in terms of root mean squared-error (RMSE) and mean absolute error (MAE). Note how though the Non-Spatial models (i.e. `ns` and `nsar1`) achieve slightly better in-sample fits, both Spatial models are more accurate at predicting the current month's surface of rent values. It is important to note how the goal in the present case is to understand the data generating process, that is, the systematic variation in the data less the random component. So while we want to minimize model error as a way to judge if our models improve on each other, we also want to smooth over sampling variability in the Craigslist data that induces local deviations compared to the true median rent. The [information criteria for the models](http://hesscl.com/smooth-sea/dashboard/index.html#model-fit){target="_blank"} generally show that the spatial specifications are more likely models of the neighborhood rent data generating process.

>>>>

---
title: "Estimating Neighborhood Rents with Scraped Data"
author: "Chris Hess and Sarah Chasins"
output: 
  html_document:
    toc: true
    toc_float: true
    includes:
      in_header: "header.html"
      after_body: "../footer.html"
abstract: "Online classified ads are a key source of information about local rental housing markets. This research first assesses the agreement of census tract rent estimates based on scraped rental advertisements with existing data sources. The second component of the paper investivates temporal and spatial dynamics in one-bedroom rents among Seattle neighborhoods using spatial models. Posterior distributions of all model hyperparameters and parameters are estimated with integrated nested Laplace approximations (INLA). Training data for models cover quarterly tract estimates for Q2 2017 through Q1 2018, while the test data are Q2 2018 median values as observed. The top model assumes that spatial and temporal dynamics are separable, i.e. independent of each other, and non-interactive. This spatial random-effects model with AR(1) temporal structure achieves a mean absolute error (MAE) of about $100 for the training data and $140 for the test data."
---

@@

Only the Craigslist and Dupre+Scott can offer insight into more recent median levels for the Seattle rental housing market. Both data suggest continued increases in the market-wide median rent for the first half of 2017, though the Craigslist slope is about twice in its size. Some of greater variation observed between the first and second data points of the Craigslist series may be related to missing listings (whether due to the early-stage process or outages), something which might bias the estimate for Q2 2017 upward in favor of regularly posted commercial listings. Though the estimates here reflect deduplicated listings, regularly updating a listing raises its likelihood of being at the top of the index our scraper works off of, thereby increasing its selection probability. This volatility in Craigslist asking rents could also be brought on by recently-developed buildings in "lease-up" adding a glut of expensive listings to the quarters' samples.

However, by the final point of Dupre+Scott's series (Fall 2017, corresponding to September), the difference compared to the Craigslist median was about $75. Mirroring other accounts about a decline in rents during Winter 2018, the Craigslist data showed a fairly precipitous decline between Q4 2017 and Q1 2018.

@@

Figure 5 shows a scatter plot/bubble plot of 2000-definition Census tract one-bedroom rent estimates between Craigslist and Dupre+Scott for the 75 tracts where comparable estimates exist. The purple dashed line indicates the loess smoother fit between the measures. Looking at the measure comparison, the first takeaway is that there is strong correlation ($r = .82$) and relative agreement between the measures (mean absolute difference = $186). The Craigslist measure reflects asking rents and the Dupre+Scott measure reflects contract rents for a neighborhood (i.e. the asking rent less an incentives a landlord offers). Given the currently low vacancy rates in most properties, the only neighborhoods where there would theoretically be more salience to the asking vs contract rent distinction (i.e. a wider gap in these indicators) are at the high end of the distribution where demand for luxury apartments has slackened some. At the high-end of the neighborhood rent distribution, however, Craigslist and Dupre+Scott are in arguably the strongest agreement.

@@

An alternative explanation for the difference between measures comes in considering the units these properties are sampling from in a given neighborhood. The greater difference between the measures among lower-rent neighborhoods (with Craigslist estimating these as more expensive than Dupre+Scott) provides some evidence for this theory. These tracts are found towards the North and South edges of the city, and generally are more residential neighborhoods where 20+ unit apartment complexes are not the entire stock of rental unit properties. In these tracts, Craigslist has a non-zero probability of sampling from any larger building but also small houses, duplex/triplexes, mother-in-law suites and recently, [Accessory Dwelling Units (ADUs)](http://www.seattle.gov/DPD/Publications/CAM/cam116a.pdf).

To the extent that units in smaller properties (in terms of the number of units) may tend to be newer, larger in housing unit square footage or offer added amenities, the true value of rent for the neighborhood will be higher than what it is estimated at using 20+ unit properties alone. Similarly, the more that smaller properties account for the overall population of rental housing units, the more likely that higher values indicated in Craigslist are not erroneous but instead reflective of a broader set of residential units.

@@

The following models learn from Craigslist data accumulated from Q1 2017 Q1 through Q1 2018 to accomplish a few different goals. The most prominent goal is to use the panel data's temporal and spatial structure to understand variation in rent levels over time and across space. A secondary goal is to forecast the Seattle neighborhood rent surface for the current quarter. Finally, the following models allow for relatively straightforward imputation of missing values based on time period and the rent values of proximate tracts.

We assume that the log of the median one-bedroom rent for Seattle neighborhoods is normally distributed and compute the posterior distributions of each model's hyperparameters and parameters $\theta_{it}$ using [integrated nested Laplace approximations (INLA)](https://folk.ntnu.no/hrue/r-inla.org/papers/inla-rss.pdf) with the default, weakly-informative priors for all random-effects. We choose to log the dependent variable because all values are positive and this transformation improves model fit.

@@

The BYM Spatial with AR(1) includes a random-effect drawn from the [Besag-York-Mollie (1991)](https://link.springer.com/article/10.1007/BF00116466) intrinsic conditional autoregressive (ICAR) distribution, along with AR(1) and exchangeable hyperparameters for the panel's temporal structure (referred to as `bym` using shorthand). The random-effect $v_i$ is assumed to come from a normal distribution whose mean is the mean of the neighbors' random effects and variance proportional to the number of neighbors (formally, $v_i|v_j, j \in ne(i) ~ \mbox{Normal}(\bar{v_i}, \frac{\sigma^{2}_v}{m_j})$. Lastly, the BYM Spatiotemporal AR(1) (`spt`) adds an exchangeable random-effect ($\delta_i \sim Normal(0, \tau_\delta)$) for each period to allow for area-specific deviations in time trends.

@@

#establish object for modeling based off of adjusting `panel`
sea_df <- panel %>% 
  mutate(actualRent = med1B, #first preserve the observed / observed-to-date
         med1B = ifelse(listingQtr == "2018 Q2", NA, med1B)) %>% #set to NA for INLA to forecast current qtr
  mutate(p1B = n1B/nListings) %>% #compute share of 1B listings
  arrange(listingQtr, GISJOIN) #arange qtr, tract
```

@@

#spatial random effect model (BYM 1991) with quarterly seasonal fixed-effect
form.bym <- log(med1B) ~ 1 + 
  f(idtract, model = "bym", #ICAR spatial RE for tract neighbors + IID RE for tract
    scale.model = T,
    graph = "../output/seatract.graph") +
  f(idqtr, model = "ar1") +
  f(idqtr1, model = "iid")

@@

#table of fit statistics
fit_stats <- sea_df %>%
  mutate(train_test = ifelse(listingQtr == "2018 Q2", "Test", "Training")) %>%
  mutate(int_err = actualRent - int_Med,
         ns_err = actualRent - ns_Med,
         nsar1_err = actualRent - nsar1_Med,
         bym_err = actualRent - bym_Med,
         spt_err = actualRent - spt_Med) %>% 
  group_by(train_test) %>%
  summarize(int_rmse = rmse(int_err[!is.na(int_err)]),
            int_mae = mae(int_err[!is.na(int_err)]),
            ns_rmse = rmse(ns_err[!is.na(ns_err)]),
            ns_mae = mae(ns_err[!is.na(ns_err)]),
            nsar1_rmse = rmse(nsar1_err[!is.na(nsar1_err)]),
            nsar1_mae = mae(nsar1_err[!is.na(nsar1_err)]),
            bym_rmse = rmse(bym_err[!is.na(bym_err)]),
            bym_mae = mae(bym_err[!is.na(bym_err)]),
            spt_rmse = rmse(spt_err[!is.na(spt_err)]),
            spt_mae = mae(spt_err[!is.na(spt_err)])) %>%
  dplyr::select(train_test, ends_with("_rmse"), ends_with("_mae"))
```

The following tables show in-sample and out-of-sample (i.e. Q2 2018) accuracy for the 5 model specifications in terms of root mean squared-error (RMSE) and mean absolute error (MAE). Note how though the Non-Spatial models (i.e. `ns` and `nsar1`) achieve slightly better in-sample fits, both Spatial models are more accurate at predicting the current month's surface of rent values. It is important to note how the goal in the present case is to understand the data generating process, that is, the systematic variation in the data less the random component. So while we want to minimize model error as a way to judge if our models improve on each other, we also want to smooth over sampling variability in the Craigslist data that induces local deviations compared to the true median rent. The [information criteria for the models](http://hesscl.com/smooth-sea/dashboard/index.html#model-fit) generally show that the spatial specifications are more likely models of the neighborhood rent data generating process.
